---
title: "Final Assignment"
output: html_document
date: "2023-12-26"
github repository: https://github.com/editor210/MY_472_Final_Assignment
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(ggpubr) #for stat_cor
library(corrplot) #for correlation table
library(plotly) #making the ggplot interactive
library(viridis) #scale color viridis
library(hrbrthemes)
library(dplyr)
library(knitr) #For printing table of Outlier Summary
library(patchwork)

```

```{r functions}
#Creating the Functions that are going to be needed on the code below


```


```{r rs_ranking, include = FALSE, echo = FALSE}

```

```{r spotify_data}

```

```{r music_brainz_data}

```

```{r popular artists_data}

```


```{r loading_data, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

artist_info_ranking <- readRDS("artist_info_ranking.rds")
top_tracks_us <- readRDS("top_tracks_us.rds")
audio_features_us <- readRDS("audio_features_us.rds")
```


```{r processing_data, include=FALSE}



```


## Introduction
Creating a ranking of the 100 greatest musical artists of all time is an ambicious task. The subjective analysis of quality and impact of an artist over history is so obvious that even [ChatGPT](https://chat.openai.com/share/65330da6-f475-4d8d-8b80-c2a4dfcbed22) aknowledges the highly individual conotation that any ranking inherently presents. Nevertheless, that is what the Rolling Stones Magazine attempted in 2010 with the publication of the ["100 Greatest Musical Artists of All Time"](https://www.rollingstone.com/music/music-lists/100-greatest-artists-147446/).

In that context, this research endeavors to explore the enduring impact of these artists and their music at the conclusion of the year 2023. Using data from Spotify, the central focus of this study revolves, first in stablishing if the ranked Rolling Stones' artists are still popular and relevant at the end of 2023 and within that scope, the study aims to decipher the factors that might contribute to the continued engagement with the music of these celebrated artists.

The Analysis will then be divided in three steps:
1 - Gathering data from Spotify to analyse how popular the ranked artists are at the end of 2023
2 - Comparing the audio features of the more popular artists ranked and compare them to see how they relate to the less popular ranked artists
3 - Comparing how the audio features of these artists correlate to the current most popular artists on the global Charts, according to the Global Weekly Spotify Charts

## Data
The ranking of ["100 Greatest Musical Artists of All Time"](https://www.rollingstone.com/music/music-lists/100-greatest-artists-147446/) was collected on the Rolling Stones website and was used as the starting point of the analysis.

The data analysis, on the other hand, was centered on information provided by the [Spotify API](https://developer.spotify.com/documentation/web-api) available information. From this source, information on artists' and tracks' popularity, features and other information was directly retrieved. The API was accessed through the [spotifyr package](https://www.rdocumentation.org/packages/spotifyr/versions/2.2.4).


## Analysis
We start this analysis by first looking at the popularity index provided by the Spotify API. According to the documentation, this variable is based on the artist's tracks popularity and the index varies between 0 and 100, with 100 being the most popular.

It is important to also notice that the popularity of each track in spotify also varies between 0 and 100, with 100 being the most popular. However, its value is based, in the most part, on the total number of plays the track has had and how recent those plays are. Although we cannot assess the specific weight that is being given to the recentness of the plays of a specific track, this measure is aligned with the main goal of this study, that is to identify the artist's relevance at the present moment in time, due to the fact that, according to the documentation, "songs that are being played a lot now will have a higher popularity than songs that were played a lot in the past".

Another limitation of the use of the variable is that it is not updated in real time, and may lag actual opularity by a few days, but since the exact moment of the analysis is not that relevant in the context of this study, this does not harm the usage of this variable.

As for the index having a range of 100, and assuming that the artists' popularity is normally distributed on Spotify, it is fair to state that all artists with a popularity over 50 could be considered "Popular" and those with a popularity lower than that threshold could be seen as "Unpopular".

On the interactive plot bellow, each dot is an artist of the Rolling Stones Ranking. The artist's Spotify popularity can be addressed on the y axis and the colors display wether the Popularity is bigger than 50 or not. The red line shows the regression line of the two variables.

```{r echo = FALSE, warning=FALSE}
# Define artist popularity based on a threshold of 50
artist_info_ranking$is_popular <- ifelse(artist_info_ranking$popularity > 50,
                                         "Popular",
                                         "Not Popular")

# Create a new dataset to use in the plot
# with additional columns and arrange the data
plot_one_data <- artist_info_ranking %>%
  # New Column with followers in thousands and round to 2 decimal places
  mutate(followers_k = round(followers.total/1000, 2)) %>%
  
  # Convert artist_spotify to a factor
  mutate(artist_spotify = factor(artist_spotify, artist_spotify)) %>%
  
  # Prepare text for tooltip in interactive plot
  mutate(text = paste("Artist: ", artist_spotify,
                      "\nPopularity on Spotify: ", popularity,
                      "\nRanking Rolling Stones: ", ranking,
                      "\nFollowers (1,000): ", followers_k, sep=""))

# Create a ggplot scatter plot with a smooth line and additional formatting
p1 <- plot_one_data %>%
  ggplot() +
  geom_point(aes(x = ranking, y = popularity, color = is_popular, text = text), size = 3, alpha = 0.5) +
  geom_line(stat = "smooth", method = lm, aes(x = ranking, y = popularity), formula = 'y ~ x', color = "red", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Rolling Stones Ranking x Spotify Popularity",
       color = NULL,
       x = "Rolling Stones Ranking",
       y = "Spotify Popularity") +
  ylim(0, 100) +
  xlim(0, 100) +
  geom_vline(xintercept = 0, color = "black", alpha = 0.1) +  # Add vertical line at x = 0
  geom_hline(yintercept = 0, color = "black", alpha = 0.1) +  # Add horizontal line at y = 0
  theme(legend.position = "bottom",
        panel.grid = element_line(color = alpha("gray", 0.2), linetype = "dashed"),  # Adjust legend position
        axis.text = element_text(size = 8),  # Adjust axis text size
        axis.title = element_text(size = 10),
        plot.title = element_text(hjust = 0.5))  # Adjust axis title size

# Convert ggplot plot to plotly for interactivity with tooltips
p1_int <- ggplotly(p1, tooltip = "text")
```

```{css, echo=FALSE, include = FALSE}
/* Centering the Plotly interactive plot

/*Source: https://stackoverflow.com/questions/47193192/r-markdown-and-plotly-fig-align-not-working-with-html-output*/

.center {
  display: table;
  margin-right: auto;
  margin-left: auto;
}
```

<div class = 'center'>
```{r, include = TRUE, echo = TRUE, fig.width = 6, fig.height = 3}
#Teste
p1_int
```
</div>

It is clear that the vast majority of the artists on the Ranking can be classified as Popular, and the red regression line also displays the negative correlation between the ranking and the popularity, which is expected since those variables are of opposite direction, meaning that a better ranking presents smaller values and a better popularity is present in bigger values. However, the small inclinement of the regression line also suggests that there is no correlation between the Ranking and the Spotify Popularity, which means that being in a better ranking position on the Rolling Stones Magazine doe not translate to a bigger popularity in 2023.

WIth that in mind, we can already stablish that the tracks of most of the artists on the Rolling Stones ranking has, indeed, endured over time, making them still popular in 2023. Now we will try to identify what differentiates the more popular artists of the Ranking from the lesser populars ones.

From the plot above and the statistical summary of these artists' popularity, it is possible to notice that the distribution is not heavily skewed, given that the median is close to the mean, the data is farily diversed distributed, without any big outliers that can be spotted based on the visual inspection of the plot.

However, the data shows a big difference from the minimum (33) to the maximum (90) popularity, which suggests that there might be differences among the artists that could potentially provide bigger or smaller popularities

```{r summary, include=TRUE, align.center = TRUE}
# Print a message
cat("Statistical Summary of Spotify Popularity\n\n")

# Print the summary of the 'popularity' variable
print(summary(artist_info_ranking$popularity))
```
The Spotify API also provides information on the Top 10 Tracks of each artist in a given market. The Rolling Stones is a North American magazine; consequently, it is anticipated that the artists featured in their rankings were, at a certain point in time, quite well-known in the United States. Therefore, a decline in popularity would carry genuine significance. In contrast, in other markets, those same artists might never have achieved fame, rendering their current popularity less meaningful in those regions.

The plot below portrays, in ascending order of popularity on the Spotify platform, the popularity of each top track by the artists featured in the Rolling Stones Ranking, along with the average track popularity and the artist's overall popularity.

```{r prep_p2, include = FALSE, fig.width=6, fig.height= 15, fig.align='center'}
# Read artist ranking information and top tracks data from RDS files
artist_info_ranking <- readRDS("artist_info_ranking.rds")
top_tracks_us <- readRDS("top_tracks_us.rds")

# Define the relative popularity based on the mean popularity threshold
artist_info_ranking$relative_pop <- ifelse(artist_info_ranking$popularity <= mean(artist_info_ranking$popularity), "Not Popular", "Popular")

# Subset the necessary columns from artist_info_ranking and top_tracks_us
subset_artist <- artist_info_ranking %>% select(id, artist_spotify, popularity, relative_pop, genres)

subset_top_tracks <- top_tracks_us %>% select(artist_id, track_popularity, track_id, track_name, album.release_date)

# Merge the two datasets based on artist IDs
merged_data_plot2 <- merge(subset_artist, subset_top_tracks, by.x = "id", by.y = "artist_id")

# Create mean track popularity data for plotting
mean_data <- merged_data_plot2 %>%
  group_by(artist_spotify) %>%
  summarise(mean_track_popularity = mean(track_popularity))

# Merge mean data with the original data
merged_data_plot2 <- merge(merged_data_plot2, mean_data, by = "artist_spotify")

# Factoring the artist_spotify for plotting in order of artist popularity
popularity_order <- artist_info_ranking %>% select(artist_spotify, popularity) %>% arrange(desc(popularity))

merged_data_plot2$artist_spotify <- factor(merged_data_plot2$artist_spotify, levels = popularity_order$artist_spotify)
```


```{r p2, include = TRUE, fig.width=6, fig.height= 15, fig.align='center'}
# Create a ggplot with three sets of points representing different popularity metrics
p2 <- ggplot(merged_data_plot2, aes(y = artist_spotify)) +
  geom_point(aes(x = track_popularity, color = "Track Popularity"), size = 1.5, alpha = 0.7) +
  geom_point(aes(x = popularity, color = "Artist Popularity"), size = 1.5, alpha = 0.7) +
  geom_point(aes(x = mean_track_popularity, color = "Average Track Popularity"), size = 1.5, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Artist and Top Tracks (US) Popularity",
       y = NULL,
       x = "Popularity",
       subtitle = "Artists are in Ascending Popularity Order") +
  scale_color_manual(
    values = c("#E63946", "#FBAF4F", "#BFD3C1"),
    name = NULL
  ) +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1),
        panel.grid = element_line(color = alpha("gray", 0.2), linetype = "solid"),
        panel.grid.major.x = element_line(linetype = "dashed"),
        panel.grid.minor.x = element_line(linetype = "dashed"),
        axis.text.y = element_text(hjust = 1, vjust = 0.5),
        plot.title = element_text(hjust = 0.5, size = 16),
        plot.subtitle = element_text(hjust = 0.5, size = 11),
        legend.position = "top") +
  geom_vline(xintercept = 0, color = "black", alpha = 0.2) +  # Add vertical line at x = 0
  scale_x_continuous(breaks = c(0, 25, 50, 75, 100),
                     expand = c(0, 0),
                     limits = c(-1, 101),
                     sec.axis = sec_axis(~., name = "Popularity"))
  
# Display the plot
p2
```

It is possible to notice that, for every artist on the Rolling Stones Ranking, the Spotify Popularity is bigger than the Average Top Tracks Popularity of the artist. However, it is also clear that if an artist's top tracks has a bigger Popularity, then, on average, the artist will also present a bigger popularity too, although the average popularity of the top tracks is never greater than the general Artist Popularity.

An interesting aspect that can also be highlighted is, when analyzing the tracks' popularity is the presence of outliers, specially among the less popular artists. Using the Interquartile Range to identify possible outliers, the table below shows that the average popularity of the artists that have a positive outlier, is below the Average Popularity of all the artists that are part of the Rolling Stones Ranking. This suggests that the most popular artists of today are not the ones with one time hits, but the ones who are able to have multiple musical hits.

```{r}
#Assign tracks that have positive outliers using IQR
merged_data_outlier <- merged_data_plot2 %>%
  group_by(artist_spotify) %>%
    mutate(
    Q1 = quantile(track_popularity, 0.25),
    Q3 = quantile(track_popularity, 0.75),
    IQR_value = Q3 - Q1,
    upper_bound = Q3 + 1.5 * IQR_value,
    outliers = ifelse(track_popularity > upper_bound, "Positive Outlier", "Not Outlier"),
    mean_popularity = mean(track_popularity),
    diff_to_mean = track_popularity - mean_popularity
  ) %>%
  ungroup()

outlier_summary <- merged_data_outlier %>%
  filter(outliers == "Positive Outlier") %>%
  summarise("Total Positive Outliers" = n(), 
            "Unique Artists With Positive Outliers" = length(unique(artist_spotify)),
             "Mean Artist Popularity" = mean(popularity),
            "Median Artist Popularity" = median(popularity),
            "Mean Difference to Average Track Popularity" = mean(diff_to_mean),
            "Max Difference to Average Track Popularity" = max(diff_to_mean))

# Convert the summary table to a nicely formatted table
kable(outlier_summary, "simple", align = 'c', caption = "Positive Outliers Summary")
```

In that sense, another characteristic that can potentially impact the current popularity of an artist would be the time. For instance, if the band does not exist anymore and they are not producing new music, the popularity of the artist might fade over time. In that sense, using the information from the Spotify API, along with data scraped from the [MusicBrainz.org](https://musicbrainz.org/) it is possible to analyse, for each of the top tracks of the artists, if, at the time that the track was released if the artist had already passed away, or if the band or group had been dissolved. 

For the plots below, it is possible to identify the tracks that were released while the artist or group was still active ("Regular Release") and the tracks that were launched after the artist or group were not performing anymore ("Release After End").
The plots are separated by artist popularity - Popular artists have more than 50 popularity points on Spotify.

```{r}
artist_info_ranking <- readRDS("artist_info_ranking.rds")
top_tracks_us <- readRDS("top_tracks_us.rds")
mbdata <- readRDS("./data/mbdata.rds")

#Selecting the necessary variables from each of the datasets
subset_mbdata <- mbdata %>% select(artist_spotify, Begin, End)
subset_artist_info_ranking <- artist_info_ranking %>% select(artist_spotify, artist_id = id, popularity)
subset_top_tracks_us <- top_tracks_us %>% select(artist_spotify, artist_id, track_name, track_id, track_popularity, album.release_date)

merge_data_p3 <- merge(subset_mbdata, subset_artist_info_ranking, by = "artist_spotify")

merge_data_p3 <- merge(merge_data_p3, subset_top_tracks_us, by = "artist_id")

# Convert dates to Date objects
merge_data_p3$Begin <- as.Date(merge_data_p3$Begin)
merge_data_p3$End <- as.Date(merge_data_p3$End)
merge_data_p3$album.release_date <- as.Date(merge_data_p3$album.release_date)

merge_data_p3$band_ended <- ifelse(merge_data_p3$End < merge_data_p3$album.release_date, "Artist not Active", "Regular Release")

merge_data_p3$band_ended[is.na(merge_data_p3$band_ended)] <- "Regular Release"

merge_data_p3$is_pop <- ifelse(merge_data_p3$popularity > 50, "Popular Artist", "Unpopular Artist")

average_popularity <- merge_data_p3 %>%
  group_by(band_ended) %>%
  summarise(avg_popularity = mean(popularity))
```

```{r}
plot_data <- merge_data_p3 %>% filter(!is.na(album.release_date))

# Create the plot
p <- ggplot(merge_data_p3, aes(x = album.release_date, y = track_popularity, color = is_pop)) +
  geom_point(size = 3, alpha = 0.7) +
  #geom_hline(data = average_popularity, aes(yintercept = avg_popularity, color = band_ended), linetype = "dashed", size = 1, show.legend = TRUE) +
  scale_x_date(date_breaks = "years", date_labels = "%Y") +
  labs(title = "Track Popularity Over Album Release Dates",
       x = "Album Release Date",
       y = "Track Popularity") +
  theme_minimal() +
  geom_smooth(method = "lm", se = FALSE)

# Display the plot
print(p)

p <- ggplot(merge_data_p3, aes(x = album.release_date, y = track_popularity, color = band_ended)) +
  geom_point(size = 3, alpha = 0.7) +
  #geom_hline(data = average_popularity, aes(yintercept = avg_popularity, color = band_ended), linetype = "dashed", size = 1, show.legend = TRUE) +
  scale_x_date(date_breaks = "years", date_labels = "%Y") +
  labs(title = "Track Popularity Over Album Release Dates",
       x = "Album Release Date",
       y = "Track Popularity") +
  theme_minimal() +
  geom_smooth(method = "lm", se = FALSE)

# Display the plot
p


```

```{r fig.width=10, message=FALSE, warning=FALSE, echo = FALSE, include = TRUE}
# Create the plot for band_ended facet
p2 <- ggplot(plot_data %>% filter(is_pop == "Popular Artist"), aes(x = album.release_date, y = track_popularity, color = band_ended)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y", limits = c(as.Date("1958-01-01"), as.Date("2023-12-12") )) +
  labs(title = "Track Popularity Over Album Release Dates",
       subtitle = "For 'Popular' Artists",
       x = "Album Release Date",
       y = "Track Popularity",
       color = "") +
  theme_minimal() +
  ylim(0,100) +
  geom_smooth(method = "lm", se = FALSE) +
  theme(legend.position = "bottom",
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_line(linetype = "dashed"),
        panel.grid.major.y = element_line(linetype = "dashed"),
        plot.title = element_text(size = 16, hjust = 0.5, color = "black", face = "bold"),
        plot.subtitle = element_text(size = 12, hjust = 0.5, color = "black")) +
  geom_hline(aes(yintercept = 0), linetype = "solid", color = "black", alpha = 0.5)

p3 <- ggplot(plot_data %>% filter(is_pop != "Popular Artist"), aes(x = album.release_date, y = track_popularity, color = band_ended)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y", limits = c(as.Date("1958-01-01"), as.Date("2023-12-12") )) +
  labs(title = "Track Popularity Over Album Release Dates",
       subtitle = "For 'Unpopular' Artists",
       x = "Album Release Date",
       y = "Track Popularity",
       color = "") +
  theme_minimal() +
  ylim(0,100) +
  #xlim(as.Date("1960-01-01"), NA) +
  geom_smooth(method = "lm", se = FALSE)+
  theme(legend.position = "bottom",
        panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_line(linetype = "dashed"),
        panel.grid.major.y = element_line(linetype = "dashed"),
        plot.title = element_text(size = 16, hjust = 0.5, color = "black", face = "bold"),
        plot.subtitle = element_text(size = 12, hjust = 0.5, color = "black")) +
  geom_hline(aes(yintercept = 0), linetype = "solid", color = "black", alpha = 0.5)

# Display the plots side by side
library(gridExtra)
grid.arrange(p3, p2, ncol = 2)

```


The first thing that can be seen in the plots is the concentration of songs released prior to 1995, but this expected given that most of the artists in the Rolling Stones Ranking are historial figures.

For the popular artists, this change is not so clear on the regression lines, so no major conclusions can be made from that.

Analysing the left side graph it is also clear the correlation of the moment of the release of the song and the status of the band. In that context, the Unpopular artists may be the ones that remained active but never achieved the same success of the past, whereas the relaunch of old hits is more likely to improve the popularity.

In that case, one interesting suggestion is that the end of the band or group, could help improve the popularity of the artist over time, maybe because Regular Released Tracks are "New Tracks" while tracks released After the End of the band are relaunches of old time successes that could have a "guarantee" better performance.It is also important to recall that the artists' popularity is derived from their music popularity, hence, the dots plotted on the graphs above can be seen as the cause of a lower popularity, and not that the smaller popularity may cause the track to be unpopular.

To finish, this study, we analyse the audio features of the tracks. Were the old songs, inherently different from the new hits? Does the song in itself makes a difference to the performance of the artist?

```{r, fig.height=20}
# Merge the dataframes based on artist_spotify
artist_info_ranking <- readRDS("artist_info_ranking.rds")
top_tracks_us <- readRDS("top_tracks_us.rds")
audio_features_us <- readRDS("audio_features_us.rds")

#First we need to merge the artist_info_ranking and the audio_feature_us to determine the popular and non popular artists
artist_info_ranking_subset <- artist_info_ranking %>% select(artist_id = id, popularity)

audio_features_us_merged <- merge(audio_features_us, artist_info_ranking_subset, by = "artist_id") 

#Creating the column is_pop in the audio_features_us_merged
audio_features_us_merged$is_pop <- ifelse(audio_features_us_merged$popularity > 50, "Popular", "Unpopular")

#Importing the data from the popular artists of 2023 audio features
popular_art_aud_feat <- readRDS("popular_artists_audio_features.rds")
#Creating a column to indicate that these are the most popular artists
popular_art_aud_feat$is_pop <- "Most Popular Today"

#Selecting the columns that will be analised
subset_cols <- c("artist_spotify",
                 "track_id",
                 "artist_id",
                 "is_pop",
                 "danceability",
                 "energy",
                 "loudness",
                 "mode",
                 "speechiness",
                 "acousticness",
                 "instrumentalness",
                 "liveness",
                 "valence",
                 "tempo")

#Creating the subsets that will be merged to be plotted
audio_features_us_subset <- audio_features_us_merged %>% select(all_of(subset_cols))
popular_art_aud_feat_subset <- popular_art_aud_feat %>% select(all_of(subset_cols))

#The dfs have the same columns, so we will just rbind them
audio_feature_plot_data <- rbind(audio_features_us_subset, popular_art_aud_feat_subset)

# Select relevant columns for comparison
comparison_cols <- c("danceability",
                 "energy",
                 "loudness",
                 "speechiness",
                 "acousticness",
                 "liveness",
                 "valence",
                 "tempo")
```


```{r fig.height=20, fig.width = 10}
# Create side-by-side box plots in a grid with 3 columns
plots <- lapply(comparison_cols, function(col) {
  ggplot(audio_feature_plot_data, aes(x = is_pop, y = get(col), fill = is_pop)) +
    geom_boxplot() +
    labs(title = paste(col),
         x = "Popularity",
         y = col) +
    theme_minimal() +
    theme(legend.position = "none")  # Remove legend for clarity
})

# Arrange plots in a grid with 3 columns
grid_plots <- wrap_plots(plots, ncol = 5)

# Display the grid of plots
print(grid_plots)
```


```{r fig.height=15, fig.width = 8}
custom_colors <- c("#1F78B4", "#FF7F00", "#33A02C")

#Plotting based on artist popularity
plots <- lapply(comparison_cols, function(col) {
  ggplot(audio_feature_plot_data, aes(x = get(col), fill = is_pop)) +
    geom_density(alpha = 0.3) +
    scale_fill_manual(values = custom_colors) +
    labs(title = paste("Density Plot of", col),
         x = col,
         fill = "Artist Popularity") +
    theme_minimal()+
    theme(legend.position = "bottom")
})
# Arrange plots in a grid with 3 columns
grid_plots <- wrap_plots(plots, ncol = 2)

# Display the grid of plots
print(grid_plots)
```





```{r, fig.height= 20, fig.width=10, eval = FALSE}
# Create scatter plots in a grid with 3 columns
plots <- lapply(comparison_cols, function(col) {
  ggplot(audio_feature_plot_data, aes(x = track_popularity, y = get(col), color = artist_popularity)) +
    geom_point(alpha = 0.3) +
    labs(title = paste(col),
         #x = "Track Popularity",
         #y = col,
         #color = "Relative Popularity") +
    )+
    theme_minimal()+
    theme(legend.position = "bottom")
})

# Arrange plots in a grid with 3 columns
grid_plots <- wrap_plots(plots, ncol = 3)

# Display the grid of plots
print(grid_plots)


```

```{r, fig.height= 20, fig.width=10, eval = FALSE}
library(ggplot2)
library(hexbin)

# Create scatter plots with hexagonal binning
plots <- lapply(scatter_cols, function(col) {
  ggplot(merged_df, aes(x = track_popularity, y = get(col))) +
    geom_hex(bins = 30) +
    scale_fill_gradient(low = "blue", high = "red") +  # Specify colors
    labs(title = paste(col),
         x = "Track Popularity",
         y = col,
         color = "Artist Popularity") +
    theme_minimal()
})

# Arrange plots in a grid with 3 columns
grid_plots <- wrap_plots(plots, ncol = 3)

# Display the grid of plots
print(grid_plots)


```


```{r, fig.height= 20, fig.width=10, eval = FALSE}
library(ggplot2)
library(hexbin)
merged_df$is_track_pop <- ifelse(merged_df$track_popularity > 50, "Popular Song", "Unpopular Song")
colnames(merged_df)
ggplot(merged_df, aes(x = track_popularity, y = get(col), group = relative_pop, fill = relative_pop)) +
    geom_density2d() +
    #scale_fill_gradient(low = "blue", high = "red") +  # Specify colors
    labs(title = paste(col),
         x = "Track Popularity",
         y = col,
         color = "Artist Popularity") +
    theme_minimal()

library(RColorBrewer)

# Define a custom color palette
custom_colors <- c("#1F78B4", "#FF7F00")

#Plotting based on track popularity
plots <- lapply(scatter_cols, function(col) {
  ggplot(merged_df, aes(x = get(col), fill = is_track_pop)) +
    geom_density(alpha = 0.3) +
    scale_fill_manual(values = custom_colors) +
    labs(title = paste("Density Plot of", col),
         x = col,
         fill = "Relative Popularity") +
    theme_minimal()+
    theme(legend.position = "bottom")
})




# Create scatter plots with hexagonal binning
plots <- lapply(scatter_cols, function(col) {
  ggplot(merged_df, aes(x = track_popularity, y = get(col))) +
    geom_hex(bins = 30) +
    scale_fill_gradient(low = "blue", high = "red") +  # Specify colors
    labs(title = paste(col),
         x = "Track Popularity",
         y = col,
         color = "Artist Popularity") +
    theme_minimal()+
    theme(legend.position = "bottom")
})

# Arrange plots in a grid with 3 columns
grid_plots <- wrap_plots(plots, ncol = 3)

# Display the grid of plots
print(grid_plots)


```

```{r, eval = FALSE}
# Subset the data frame to include only the relevant columns
subset_df <- merged_df[c("danceability", "energy", "key", "loudness", "mode", "speechiness", "acousticness", "instrumentalness", "liveness", "valence", "tempo", "artist_popularity")]

# Assuming your model was created using lm function
model <- lm(artist_popularity.x ~ ., data = subset_df)

# Assuming diagnostic_plots function is defined as before
diagnostic_plots <- function(model) {
  # ... (unchanged code)
}

# Call the function with your regression model
diagnostic_plots(model)


```
```{r, eval = FALSE}
# Install and load necessary packages (if not already installed)
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  install.packages("ggplot2")
}
if (!requireNamespace("cowplot", quietly = TRUE)) {
  install.packages("cowplot")
}
library(ggplot2)
library(cowplot)

# Define diagnostic_plots function
diagnostic_plots <- function(data, dependent_var, independent_vars) {
  # Fit a multiple linear regression model
  model <- lm(paste(dependent_var, "~", paste(independent_vars, collapse = " + ")), data = data)

  # Residuals vs Fitted
  plot1 <- ggplot(model, aes(fitted.values, residuals)) +
    geom_point() +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    labs(title = "Residuals vs Fitted", x = "Fitted values", y = "Residuals")

  # Normal Q-Q Plot
  plot2 <- ggplot(model, aes(sample = residuals)) +
    stat_qq() +
    stat_qq_line() +
    labs(title = "Normal Q-Q Plot", x = "Theoretical Quantiles", y = "Sample Quantiles")

  # Scale-Location (Square Root of Standardized Residuals) vs Fitted
  plot3 <- ggplot(model, aes(fitted.values, sqrt(abs(residuals)))) +
    geom_point() +
    geom_smooth(se = FALSE, method = "loess", formula = y ~ x) +
    labs(title = "Scale-Location Plot", x = "Fitted values", y = "Square Root of Standardized Residuals")

  # Residuals vs Leverage
  plot4 <- ggplot(model, aes(hatvalues, residuals)) +
    geom_point() +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    labs(title = "Residuals vs Leverage", x = "Leverage", y = "Residuals")

  # Combine plots into a grid
  plot_grid(plot1, plot2, plot3, plot4, ncol = 2, labels = "AUTO")
}

# Call the function with your data frame and variable names
diagnostic_plots(subset_df, "artist_popularity", c("danceability", "energy", "key", "loudness", "mode", "speechiness", "acousticness", "instrumentalness", "liveness", "valence", "tempo"))


```

Introduction: A brief introduction to the research question and your approach to answering it. You do not need to cite any literature or write a literature review.

Data: A discussion of the data sources you used, how you accessed them, how you processed the data, the structure of your final analysis dataset(s), and so on.

Analysis: A presentation of your analysis, including figures/graphs/maps, and a discussion of your findings. In general, we do not expect you to conduct or interpret any formal statistical tests, though you may do this if you wish. Remember that your discussion should translate your specific analysis and results back to the level of the research question.

Code Appendix: As in the previous assignments, all code that you do not wish to directly include in your report should be included in a code appendix at the end of the document.

- Parliament and Funkadelic are two bands that only recently have been merged to one, so the decision was made to add both of them to the list of artists, with the same ranking
- 

#Next steps
# 1 - After I finish tidying the artist_id_pop data I add the id to the previous rs_ranking
# Create a relational SQL databse to get the information using the spotify id as a key
# 2 - Get top tracks for each artist
# 3 - Get audio features for every track
# 4 - Find similarities and graphs to show how the artists popularity and genres and audiofeatures relate
# 5 - Get the top 50 global artists in Spotify
# 6 - See how the top 50 audio features is related to the artists in the ranking
# 7 - Argue that the popularity is a controversial thing and can change from country to country
# 8 - The ranking is 100% anglophone (plot a map of the artists origin)
# 9 - Get the end date of each band or artist from the other website to see if the end date is related to the popularity
# 10 - Discuss the results
